---
title: "Differentiation Assay Script"
author: "Maxime"
date: "2023-02-17"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(flowCore)
library(ggcyto)
library(openCyto)
library(gtools)
library(patchwork)
library(purrr)
library(stringr)
library(dplyr)
```
## Notes before reading this document
- for {golem}: Inside your Vignettes folder: call usethis::use_vignette() to create the skeleton for a Vignette, which will then be used as package documentation. Use this .Rmd document as the package vignette.

- one analysis is not yet described/implemented in this document: binning GFP-positive cells. I want to show a specific example in order to clarify the idea. I don't think this is a lot of work, as the general idea is similar to what I do here.

- the parameters for my ggcyto plot can be set once with **settings <- ggcyto::ggcyto_par_set()** so that size of the codebase is reduced. 

- I think it's best to split this .Rmd file into multiple ones, one for each part of the end application (https://engineering-shiny.org/building-ispum-app.html) (import, curation, gating... these will essentially also be the "modules" of the app, if I understood the module concept from {golem} correctly)

## Loading dependencies
Visualization and analysis of flow cytometry data requires some R packages. The code below ensures that all the necessary packages and dependencies are installed and loaded. 

## Preliminary steps

### Output folder
During our analysis, we will generate many different plots and tables. We therefore first need to create a folder that will contain all our output. We will name this folder "data.output". The code below checks if such a folder already exists in our current working directory and if not, creates it. In case it already exists, you will see the message "FALSE"appearing in your console. 

``` {r}
if (!fs::dir_exists("data_output")) fs::dir_create("data_output")
```

### FCS file folder
FCS files exported from your FACS machine commonly contain multiple datasets. Each dataset contains data from the analysis of a single well (i.e if you analyze a 96-well plate, your exported FCS file contains 96 datasets). The flowCore package we use here provides two data structures that provide a means for analyzing our data in R:

* flowFrame: stores information about single FACS runs (i.e single dataset)
* flowSet: stores information from multiple FACS runs (i.e multiple datasets)

Since our FACS machine exports a single FCS file with multiple datasets, we first need to create individual FCS files from each dataset that will later be read into a *flowSet*. The individual FCS files are saved to a folder named "fcs.files". The code below creates such a folder if not yet present in our current working directory. 

```{r files folder}
if (!fs::dir_exists("fcs_input")) fs::dir_create("fcs_input")
```

### FCS file name
We now need to tell R the name of the FCS file exported from our FACS machine. In our case, this is a single file containing multiple datasets. The code below lists files from current working directory with an ".fcs" extension and assigns their name to a variable called *filename*. 

Two important considerations:

1. You can always check the current working directory by typing getwd() in the console and pressing "Enter".  

2. Make sure that your current working directory only contains **a single** FCS file. Otherwise, the code below will assign a vector of multiple filenames to the *filename* variable, and subsequent lines of code will procude an error.  works if a **single** FCS file is present in the current working directory.

```{r}
filename <- fs::dir_ls(glob = "*.fcs")
filename
```

### Number of datasets in our FCS file
To determine the number of necessary iterations necessary to create FCS files from  datasets, we need to know how many datasets our initial FCS file contains. The code block below finds the number of datasets in our initial FCS file and assigns it to the variable *n.datasets*. 

```{r}

n_datasets <- function(filename) {
	# Adapted code from https://github.com/RGLab/flowCore/blob/ba3b6ffed5310c1c0618487ab163c0142d8cab8f/R/IO.R
	con <- file(filename, open = "rb")
	
	offsets <- flowCore:::readFCSheader(con)
	offsets <- matrix(offsets, nrow = 1, dimnames = list(NULL, names(offsets)))
	txt <- flowCore:::readFCStext(con, offsets[1, ],emptyValue = FALSE)
	
	addOff <- 0
	
	if ("$NEXTDATA" %in% names(txt)) {
		nd <- as.numeric(txt[["$NEXTDATA"]])
	} else
		nd <- 0
	
	txt.list <- list(txt)
	i <- 1
	while (nd != 0) {
		i <- i + 1
		addOff <- addOff + nd
		offsets <- rbind(offsets, flowCore:::readFCSheader(con, addOff))
		this.txt <- flowCore:::readFCStext(con, offsets[nrow(offsets),], emptyValue = FALSE)
		nd <- as.numeric(this.txt[["$NEXTDATA"]])
		txt.list[[i]] <- this.txt
	}
	length(txt.list)
}
#nb_ds <- n_datasets("2023-02-06_at_07-14-51pm cmpd test ANALYSED.fcs")
nb_ds <- n_datasets(filename)
nb_ds
```

After running this code block, the number of datasets will be printed to the console. You should always check if this number corresponds to the number of analyzed wells and you should not proceed if these two numbers do not match.  

### Reading datasets as individual FCS files into R
We will now read each individual dataset of our initial FCS file and convert it to an individual FCS file saved into our fcs.files folder. After running this code block, you can check if this folder indeed contains our FCS files.

```{r}
# define the path of the folder our FCS files should populate
# must be fcs_input

walk(seq_len(nb_ds), \(x) {
	fr <- read.FCS(filename,
								 dataset = x,
								 transformation = FALSE,
								 truncate_max_range = FALSE,
								 alter.names = TRUE,
								 emptyValue = FALSE)
	message(paste("Write file #", x, "well", fr@description$`$WELLID`))
	# write the individual flowframe objects to individual FCS files
	write.FCS(fr, fs::path("fcs_input", glue::glue("dataset_{fr@description$`$WELLID`}.fcs")))
}) 

```

We now read these individual FCS files into a **flowSet**, flowCore data structure that will allow us to perform our analysis. 

```{r}
fs <- read.flowSet(fs::dir_ls("fcs_input", glob = "*.fcs"),
									 truncate_max_range = FALSE,
									 alter.names = TRUE,
									 transformation = FALSE)
fs
```

You can click on the fs variable in the R environment to inspect the flowSet we've created. A flowSet consists of different flowFrames, each flowFrame collecting data from a single sample analyzed on your FACS machine. In case we need it later, we add the well id to the phenotypic data (pData) of our flowset using the code below.

```{r}
pData(fs)$well <- str_extract(pData(fs)$name, "[A-Z]\\d{2}")
```


## Compensation

## Gating set
Gating is an essential part during the analysis of flow cytometry data. The flowWorkSpace package provides a data structure called **GatingSet** that facilitates interaction with gated data. We therefore transform our flowSet to a gatingSet using the code below.

```{r}
gs <- GatingSet(fs)
```

## Gating your data
### Gating intact cells
Preparing cells for FACS analysis (fixation, permeabilization, etc.) can lead to cell death and therefore debris in our sample that that should be excluded from our analysis. Since intact cells have a different morphology and size compared to debris, we can distinguish them in a SSC vs FSC plot. 

```{r, out.width="100%", message=FALSE}
ggcyto(gs[[1:3]], aes(x = SSC.HLin, y = FSC.HLin), subset = "root") +
  geom_hex(bins = 150) +
  ggcyto_par_set(limits = "instrument") +
  theme(complete = TRUE,
        axis.text.x = element_text(angle = 45, size = 10)) +
  theme_bw()
```

In SSC vs FSC plots, dead cells and debris cluster in the bottom left corner because of their morphology. However, some variation across samples can be observed regarding the shape and size of this cluster. To properly define a gate excluding this region we think it's best to add a debris control (staurosporin treatment, concentration to be determined) to your samples.

Below we define a polygonal gate that to excludes most of the cells clustering in the lower left corner in the previous plot. We summarize the x- and y-coordinates of polygon corner a 2-column matrix. The column names of the matrix have to be set to SSC and FSC. The filterId argument of the polygonGate() function can be used to name the gate. Later, this will allows us to refer to the gate using this name.

```{r}
#defining the polygon gate matrix
pgn.cut <- matrix(c(0, 12500, 99000, 99000,0,
										12500, 0, 0, 99000, 99000),
									ncol = 2,
									nrow = 5)
#setting the correct column names
colnames(pgn.cut) <- c("SSC.HLin","FSC.HLin")

#creating the gate using polygonGate()
non.debris.gate <- polygonGate(filterId = "NonDebris", .gate = pgn.cut)
```

Before adding the gate to our gatingSet object we should visualize it.

```{r, out.width = "100%"}
ggcyto(gs[[1]], aes(x = SSC, y = FSC), subset = "root") +
  geom_hex(bins = 200) +
  geom_gate(non.debris.gate) +
  ggcyto_par_set(limits = "instrument") +
  geom_stats(type = c("percent", "count"), location = "gate", adjust = c(0.15,0.85))+
  theme(complete = TRUE,
        axis.text.x = element_text(angle = 45)) +
  theme_bw()
```
We can now add the gate to our gatingSet object. The recompute() function performs the actual filtering operation on our data. In addition, we plot the gating tree.

```{r}
# add the gate to the GatingSet
gs_pop_add(gs, non.debris.gate, parent = "root")

# recompute the GatingSet
recompute(gs)

# plot the GatingSet to verify the underlying hierarchy
plot(gs)
```

### Gating GFP-positive cells
We want to focus our analysis on cells successfully transfected with GFP-tagged KRAS. Random noise as well as autofluorescence from untransfected cells inevitably leads to some degree of "background" signal in the GFP channel. We can determine the threshold for background signals using our unstained and MyoHC control. Let's plot the GFP intensitiy distribution for these two samples.

```{r, out.width="100%"}
ggcyto(gs[[c(1,3)]], aes(x = "GRN.B.HLin"), subset = "NonDebris") +
  geom_density(fill = "palegreen") +
  ggcyto_par_set(limits = "instrument") +
  scale_x_flowjo_biexp() +
  theme_bw()
```

Although the differences are small, these two distributions are not identical, indicating that background intenities are fluctuating across samples and a threshold intensity for GFP-positive cells should therefore consider both distributions. 
Conceptually, we apply the quantile_gate() function that determines a cutpoint by the events quantile. In our case, we set the *probs* argument to 0.99 to find a cutpoint that encompasses 99% of the data. Applying this function to both negative GFP controls and averaging the result, we obtain a cutpoint for our final gate. 

**Convention. Controls should be in a precise order: (1) unstained, (2) GFP only, (3) MYHC only.**

```{r}
#extract NonDebris population data and change to flowSet
nonDebris_data <- gs_pop_get_data(gs[[c(1,3)]], "NonDebris") |> cytoset_to_flowSet()

# create a quantileGate for both samples
gfp_test_gate <- fsApply(nonDebris_data, 
                         function(fr) openCyto::gate_quantile(fr, 
                                                              channel = "GRN.B.HLin",
                                                              probs = 0.99))

# average the lower threshold from both gates
lower_limit_gfp_gate <- mean(c(gfp_test_gate[[1]]@min, gfp_test_gate[[2]]@min))

# create the final gate
gfp_gate <- rectangleGate("GRN.B.HLin" = c(lower_limit_gfp_gate, Inf), 
                          filterId = "GFP+")
```

We can plot our result...

```{r, out.width="100%"}
ggcyto(gs, aes(x = "GRN.B.HLin"), subset = "NonDebris") +
  geom_density(fill = "palegreen") +
  geom_gate(gfp_gate) +
  ggcyto_par_set(limits = "instrument") +
  scale_x_flowjo_biexp() +
  theme_bw()
```

... and add the gate to our gatingSet object, plot the gating tree and recompute the gatingSet. 

When adding the gate to the gatingSet, we need to specify the parent population. In our case, this should be the "NonDebris".

```{r}
# add the gate to the GatingSet
gs_pop_add(gs,gfp_gate, parent = "NonDebris")

# plot the GatingSet to verify the underlying hierarchy
plot(gs)

#recompute the Gating Set
recompute(gs)
```

### Identify and select MYHC-positive cells
We apply the same logic to gate for myosin-positive cells.

```{r, out.width="100%"}
ggcyto(gs[[c(1,2)]], aes(x = "RED.R.HLin"), subset = "NonDebris") +
  geom_density(fill = "pink") +
  ggcyto_par_set(limits = "instrument") +
  scale_x_flowjo_biexp() +
  theme_bw()
```

We retrieve data from the NonDebris population, create a flowSet from a cytoSet and apply a quantile gate to obtain a threshold value that excludes 99% of the cells. 

```{r}
nonDebris_data <- gs_pop_get_data(gs[[c(1,2)]], "NonDebris") |> cytoset_to_flowSet()

myo_test_gate <- fsApply(nonDebris_data, 
                         function(fr) openCyto::gate_quantile(fr, 
                                                              channel = "RED.R.HLin", 
                                                              probs = 0.99))
# average the lower threshold from both gates
lower_limit_myo_gate <- mean(c(myo_test_gate[[1]]@min, myo_test_gate[[2]]@min))

# create the final gate
myo_gate <- rectangleGate("RED.R.HLin" = c(lower_limit_myo_gate, Inf), 
                          filterId = "MYO+")
```

We can check the gate in a plot...

```{r, out.width="100%", warning=FALSE}
ggcyto(gs[[c(1,2)]],aes(x = "RED.R.HLin"), subset = "NonDebris") +
  geom_density(fill = "pink") +
  geom_gate(myo_gate) +
  ggcyto_par_set(limits = "instrument") +
  scale_x_flowjo_biexp() +
  theme_bw()
```

...add the gate to the gatingSet, plot the gating tree and recompute the gatingSet. When adding the gate to the gatingSet, we need to specify the parent population. Now this should correspond to "GFP+". 

```{r}
#add gate to GFP+ population
gs_pop_add(gs,myo_gate, parent = "GFP+")

#plot the GatingSet
plot(gs)

#recompute
recompute(gs)
```

### Binning GFP+ cells
We want to subset the MYO+ population into three distinct subpopulations based on GFP expression:

- GFP-low: intensities = [lower_limit_gfp_gate,100]
- GFP-medium: intensities = [101, 350]
- GFP-high: intensities = [351, 1000]

```{r}
gate_limits <- list(low = list(c(lower_limit_gfp_gate, 100)),
                    medium = list(c(101, 350)),
                    high = list(c(351,1000)))

filter_names <- c("GFP-low", "GFP-medium", "GFP-high")

# why is this working precisely?
gates <- lapply(gate_limits, function(x) {
  names(x) <- "GRN.B.HLIN"
  rectangleGate(x)
})


gfp_low <- rectangleGate("GRN.B.HLin" = c(lower_limit_gfp_gate, 100), filterId = "GFP-low")
gfp_medium <- rectangleGate("GRN.B.HLin" = c(101, 350), filterId = "GFP-medium")
gfp_high <- rectangleGate("GRN.B.HLin" = c(351, 1000), filterId = "GFP-high")

gs_pop_add(gs, gfp_low, parent = "MYO+")
gs_pop_add(gs, gfp_medium, parent = "MYO+")
gs_pop_add(gs, gfp_high, parent = "MYO+")

plot(gs)

recompute(gs)
```

We can visualize the gates.

```{r}
ggcyto(gs[[5:10]], aes(x = "GRN.B.HLin"), subset = "MYO+") +
  geom_density(fill = "forestgreen") +
  scale_x_flowjo_biexp() +
  theme_bw() +
  geom_gate(gfp_low) +
  geom_gate(gfp_medium) +
  geom_gate(gfp_high)
```

Our control samples have served their purpose to narrow down the population we want to perform our analysis on. As shown below, the GFP-low/medium/high populations only contain, if any at all, very few cells. Before we proceed our analysis, we should therefore remove these three control samples from the gaingSet (reasons explained below).

```{r}
# inspect data
# head(data.frame(gs_pop_get_count_fast(gs)))
# 
# # remove control samples from gatingSet
# 
# gs <- gs[[4:length(gs)]]
```

### Myosin expression-based clustering

We're interested in myosin expression in our GFP-low/medium/high subpopulations. Usually we observe two distinct peaks: one with high and one with low myosin expression. Our goal is to automate the gating procedure and therefore we need an algorithm that automatically finds the cutpoint between these two peaks based on the underlying data.

**Note: we run into problems for samples that do not contain a sufficient amount of cells. Somehow the clustering algorithm cannot perform necessary computations. We therefore either need a way of removing these samples (for this we would need to find a way to know how many observations are sufficient) or we file an issue for the non-working mindensity function so that the developers can suggest a solution?**

```{r}
# extract the data
data_gfp_low <- gs_pop_get_data(gs, y = "GFP-low") |> cytoset_to_flowSet()
data_gfp_medium <- gs_pop_get_data(gs, y = "GFP-medium") |> cytoset_to_flowSet()
data_gfp_high <- gs_pop_get_data(gs, y = "GFP-high") |> cytoset_to_flowSet()


# apply a clustering algorithm
gfp_low_myo_high <- fsApply(data_gfp_low, 
                            function(fr) gate_flowclust_1d(fr,params = "RED.R.HLin",
                                                           K = 2, 
                                                           cutpoint_method = "min_density",
                                                           filterId = "GFP-low-MYO-high"))


gfp_medium_myo_high <- fsApply(data_gfp_medium, 
                            function(fr) gate_flowclust_1d(fr,params = "RED.R.HLin",
                                                           K = 2, 
                                                           cutpoint_method = "min_density",
                                                           filterId = "GFP-medium-MYO-high"))

gfp_high_myo_high <- fsApply(data_gfp_high, 
                            function(fr) gate_flowclust_1d(fr,params = "RED.R.HLin",
                                                           K = 2, 
                                                           cutpoint_method = "min_density",
                                                           filterId = "GFP-high-MYO-high"))

test_function <- function(fr) {
  return(tryCatch(gate_flowclust_1d(fr,params = "RED.R.HLin",K = 2, cutpoint_method = "min_density", filterId = "GFP-high-MYO-high"), error = function(e) NULL))
  }

gfp_low_myo_high <- fsApply(data_gfp_low, test_function)
gfp_medium_myo_high <- fsApply(data_gfp_medium, test_function)
gfp_high_myo_high <- fsApply(data_gfp_high, test_function)

gfp_low_myo_high <- gfp_low_myo_high[-which(sapply(gfp_low_myo_high, is.null))]
gfp_medium_myo_high <- gfp_medium_myo_high[-which(sapply(gfp_medium_myo_high, is.null))]
gfp_high_myo_high <- gfp_high_myo_high[-which(sapply(gfp_high_myo_high, is.null))]
```

```{r, warning=FALSE}

ggcyto(gs[[5:8]], aes(x = "RED.R.HLin"), subset = "GFP-medium") +
  geom_density() +
  scale_x_flowjo_biexp() +
  theme_bw() +
  geom_gate(gfp_high_myo_high)

ggcyto(gs[[5:8]], aes(x = "RED.R.HLin"), subset = "GFP-medium") +
  geom_density() +
  scale_x_flowjo_biexp() +
  theme_bw() +
  geom_gate(gfp_medium_myo_high)
```


**below: out-of-data, review**
We add the gate to the gatingSet, plot the gating tree and recompute the gatingSet.
```{r}
# add the gate to the GatingSet
gs_pop_add(gs, myosin.flowclust.gate, parent = "MYO+")

# plot the GatingSet to verify the underlying hierarchy
plot(gs)

#recompute the Gating Set
recompute(gs)
```

Let's visualize the final gate with the corresponding population statistics.

```{r, warning=FALSE, out.width="100%"}
ggcyto(gs[[5:15]], aes(x = GFP, y = APC), subset = "MYO+") +
  geom_point(size = 0.05, alpha = 0.5) +
  geom_gate(myosin.flowclust.gate) +
  scale_x_flowjo_biexp() +
  scale_y_flowjo_biexp() +
  ggcyto_par_set(limits = "instrument") +
  geom_stats()+
  theme(axis.text.x = element_text(angle = 45))+
  theme_bw()
```

We finally retrieve the population statistics.

```{r}
# get stats of the gated population
myo.positive.count <- data.frame(gs_pop_get_count_fast(gs, statistic = c("count")))
myo.positive.stats <- myo.positive.count %>% mutate("Freq" = Count/ParentCount) %>% select(1,3,5,2,4,6)
view(head(myo.positive.stats))
```

We can filter the data to only contain the count and percentage of myosin-high cells in our myosin-positive population.

```{r, out.width="75%"}
myo.positive.stats.filtered <- myo.positive.stats %>% 
  dplyr::filter(Population == "/NonDebris/GFP+/MYO+/MYO-high")

wells <- str_extract(myo.positive.stats.filtered$name, "[A-Z][0-9]+")

myo.positive.stats.filtered <- myo.positive.stats.filtered %>% 
  mutate(well = wells)

print(myo.positive.stats.filtered)
```

If we want, we can plot these results.

```{r, out.width="100%"}
#plot results with geom_col()
ggplot(myo.positive.stats.filtered, aes(x = well, y = Freq))+ 
  geom_col()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 45, 
                                   face = "bold", 
                                   size = 14, 
                                   vjust = 0.7),
        axis.ticks = element_blank())+
  ylab(label = "MYO-high cells (% of MYO+ cells)")+
  xlab(label = "")+
  scale_y_continuous(expand=c(0,0), limits = c(0,max(myo.positive.stats.filtered$Freq)+0.1))
```
