---
title: "Data Curation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Curation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Gating live cells
Preparing samples for flow cytometry inevitably lead to the formation of damaged or dead cells (also called "debris") that need to be excluded from the analysis. Conveniently, we can distinguish such cells in a SSC vs FSC plot and remove them with appropriate gates.

In flowFate, we remove debris with a "hard" gate (FIGURE!!!). This means that the gate boundaries are **predefined** and users are not able to manually change them. This needs to be considered during acquisition to ensure that intact cells accumulate in this pre-defined window. 

**Note:** Advanced R users could nevertheless manually adjust these gate boundaries by changing the respective gating matrix in the source code.

### 1) Setting up the gate matrix
We first need to set up a m x n gate matrix:

* **n**, the number of columns, corresponds to the gating dimensions (here n = 2 since cells are gated based on their location in a 2D SSC vs FSC plot)

* **m**, the number of rows, corresponds to the number of individual points that make up the rectangle gate. 

Each row should then contain the X and Y coordinates (in the first and second column of the matrix, respectively) of the latter points. We then rename the columns to match the gating dimensions. Since the FSC and SSC channels change depending on the flow cytometer, the user can select the appropriate channel names and the columns of the gate matrix are adapted based on this selection.

```{r}
# creating the gate matrix
pgn_cut <- matrix(c(12500, 99000, 99000,0,0,6250, 6250, 99000, 99000,12500), ncol = 2, nrow = 5)
# renaming the columns to match the gating dimensions
colnames(pgn_cut) <- c("Side scatter", "Forward scatter")
```

```{r, echo=FALSE}
pgn_cut
```


### 2) Create the gate
We create the gate using flowCore's `polygonGate( )` function. The filterId argument uniquely identifies the generated gate. 

```{r, eval = FALSE}
gate_non_debris <- reactive(polygonGate(filterId = "NonDebris", .gate = pgn_cut))
```


### 3) Apply the gate to the GatingSet
Up to now, we have only created a gate from a gate matrix. To **add** this gate to the GatingSet (`r$gs`, created during data import (reference to import vignette and section)), we use `gs_pop_add( )`. Recomputing `r$gs` with `recompute( )`, we **apply** the gate to our data.

```{r, eval = FALSE}
gs_pop_add(r$gs, gate_non_debris(), parent = "root")
recompute(r$gs)
```


## Gating GFP+ cells
flowFate was mainly developed to study the effect transfection of GFP-labeled KRas constructs has on C2C12 differentiation. Our analyis should therefore be focused on transfected cells only. Untransfected cells however, due to background noise and autofluorescence, can also emit fluorescence captured in the green channel of our flow cytometer. To distinguish transfected from untransfected cells, we apply a GFP threshold to our data so that cells below a certain threshold intensity are removed from further analysis.

To determine the value of this threshold, we use:

* a **double-negative control** sample: untransfected, unlabelled C2C12

* a **MyHC-only control** sample: untransfected C2C12, labelled with an antibody against myosin heavy-chain (MyHC)

For both control samples, the observed GFP signal (green channel of the flow cytometer) is unspecific and is illustrated below.

SCREENSHOT

The key idea here is to determine a GFP threshold for which most of the cells in these two controls samples would be below this threshold. We then use this threshold to create a gate that only keeps cells if they are **above** this threshold, i.e the GFP signal is specific.

### 1) Determine a GFP threshold intensity
Most of the work here is wrapped into a custom function `get_lowerLimit( )`.

```{r, eval=FALSE}
get_lowerLimit <- function(gs, datasets, node, ch_gate, r) {
  require(flowCore)
  # extract the data from the two control samples as a flowSet
  x <- gs_pop_get_data(r$gs[[datasets]], node) |> cytoset_to_flowSet()
  
  # for each dataset of the extracted flowSet, create a quantile gate using the correct channel name (in our case this will be the name of the GFP channel selected by the user)
  y <- fsApply(x, function(fr) {
    openCyto::gate_quantile(fr, channel = ch_gate, probs = 0.99)
  })
  
  # return the average of two quantile gate's lower limit
  z <- mean(c(y[[1]]@min, y[[2]]@min))
}
```

```{r, eval=FALSE}
lower_limit_gfp_gate <- reactive({
  get_lowerLimit(gs = r$gs, datasets = c(input$negative_control, input$positive_control_myhc), node = "NonDebris", ch_gate = input$kras_channel, r = r)
}) |> bindEvent(input$Curate)
```


```{r, eval=FALSE}
create_quantile_gate <- function(samples, gate_channel) {
  require(flowCore)
  fsApply(samples, function(fr) {
    print(fr)
    openCyto::gate_quantile(fr, channel = gate_channel, probs = 0.99)
  })
}
```

