---
title: "01 Import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data-import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Our analysis starts with data import. While some flow cytometry acquisition software export data from different samples as **separate .fcs files**, others export a **single, merged .fcs file** containing data from multiple samples. The $NXTDATA keyword can be used to retrieve data from individual samples. 

In flowFate, users can upload a merged file or individual files. If individual files are uploaded, they must be grouped inside a folder.

**Note:** Imported files must be FCS3.0. 

## A - Merged file upload
### 1) Fetch the number of datasets
We first fetch the number of datasets contained in the merged .fcs file (code adapted from [RGLab](https://github.com/RGLab/flowCore/blob/ba3b6ffed5310c1c0618487ab163c0142d8cab8f/R/IO.R)).


```{r, eval = FALSE}
n_datasets <- function(filename) {
  # Adapted code from https://github.com/RGLab/flowCore/blob/ba3b6ffed5310c1c0618487ab163c0142d8cab8f/R/IO.R

  # the keyword $NEXTDATA contains either a zero when there are no next data
  # or a positive integer for the next dataset
  counter <- 0
  nextdata <- 1
  while (nextdata != 0) {
    counter <- counter + 1
    nextdata <- read.FCSheader(filename, keyword = "$NEXTDATA", emptyValue = FALSE, dataset = counter)
    nextdata <- as.integer(nextdata[[1]]["$NEXTDATA"])
  }
  counter
}
```


### 2) Write individual .fcs files
We write the individual datasets inside the merged .fcs file as individual .fcs files to a temporary folder using the custom `split_1_fcs( )` function. We append the $WELLID keyword to the filename.

```{r, eval = FALSE}
split_1_fcs <- function(nb, input_file) {
  # create a folder named fcs_input in a temporary location
  if (!dir_exists(path(path_dir(input_file), "fcs_input"))) dir_create(path(path_dir(input_file), "fcs_input"))
  
  # walk along every dataset and read it as a flowFrame (data structure from flowCore)
  walk(seq_len(nb_ds), \(x) {
    fr <- read.FCS(filename, dataset = x,
                   transformation = FALSE,
                   truncate_max_range = FALSE,
                   alter.names = TRUE,
                   emptyValue = FALSE)
    message(paste("Write file #", x, "well", fr@description$`$WELLID`))
    
    # write flowframes to individual FCS files inside the fcs_input folder
    write.FCS(fr, fs::path("fcs_input", paste0("dataset_", fr@description$`$WELLID`, ".fcs")))
  })
  
  # return the complete path of the fcs_input folder
  path(path_dir(input_file), "fcs_input")
}

# create a reactive expression that, when called, executes the split_1_fcs() function and returns a directory
individual_fcs <- reactive(split_1_fcs(nb_ds(), input$filename$datapath))
```

### 3) Create a flowSet
We read the individual .fcs files (stored in the fcs_input temporary folder) into a flowSet, a data structure from the flowCore package). The reactive expression `individual_fcs()` returns the directory of fcs_input temporary folder.

```{r, eval = FALSE}
fs <- reactive(read.flowSet(fs::dir_ls(individual_fcs(), glob = "*.fcs"), 
                            truncate_max_range = FALSE,
                            alter.names = TRUE,
                            transformation = FALSE))
```

## B - Individual file upload

**Note: **Individual files (all FCS3.0) have to be grouped inside a folder. Steps 1) and 2) from above are now dispensable and we can start by creating a flowSet.

### 1) Create a flowSet
```{r, eval = FALSE}
fs <- reactive({
  read.flowSet(path = filename(),
               truncate_max_range = FALSE,
               alter.names = TRUE,
               transformation = FALSE,
               emptyValue = FALSE)
})
```

**Note:** `filename()` is a reactive expression capturing the patch of the folder a user uploads.

## C - Create a GatingSet
The flowFrame and flowSet data structures from [flowCore](https://bioconductor.org/packages/release/bioc/html/flowCore.html) are very useful for handling flow cytometry data in R. The GatingSet data structure from [flowWorkspace](https://www.bioconductor.org/packages/release/bioc/html/flowWorkspace.html) is, however, better suited for storing and manipulating gated data.

### 1) Create a GatingSet from a flowSet
We use the `GatingSet()` function from [flowWorkspace](https://www.bioconductor.org/packages/release/bioc/html/flowWorkspace.html) to create a GatingSet from the previous flowSet.

```{r, eval = FALSE}
gs <- reactive({GatingSet(fs())})
```


## D - Sharing reactive values across modules
The [{golem}](https://thinkr-open.github.io/golem/) framework promotes the use of modules to split a Shiny app into separate, smaller parts. Different [techniques](https://engineering-shiny.org/common-app-caveats.html?q=sharing%20data%20#a.-sharing-data-across-modules) are available in {golem} to allow communication between modules. 
We opted for the ["stratégie du petit r"](https://engineering-shiny.org/structuring-project.html?q=strat#b.-the-strat%C3%A9gie-du-petit-r). Briefly, a list of reactive values - called ```r``` - is passed along modules. 

### 1) Completing the reactive list ```r``` 
We add the GatingSet, flowSet and nb_ds (number of datasets) reactive expression to the reactive list. These variables can now be used in other modules by calling ```r$gs```, ```r$fs``` or ```r$nb_ds```. To read reactive expressions in Shiny, they have to be wrapped in inside `observe( )`.

```{r, eval = FALSE}
    observe({
      r$gs <- GatingSet(fs())
      r$fs <- fs()
      r$nb_ds <- nb_ds()
    })
```
