---
title: "04 Export"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Our data analysis process generated different populations: 

-   the NonDebris population
-   the GFP+ population
-   the GFP-low, GFP-medium, GFP-high population
-   the MyHC+

Each population is characterized by its size i.e the number of cells that met the gating criteria for this population. This information is saved as a population table which can be retrieved using the `gs_pop_get_stats( )` function. We wrap this function into a `map_df( )` call to extract and combine the population tables from each dataset in the GatingSet to a large and unique population table.

```{r, eval=FALSE}
reactive({
  purrr::map_df(r$gs, \(x) as.data.frame(gs_pop_get_stats(x)))
})
```

Users are able to download the population table. We implemented this using Shiny's `downloadHandler( )` function as well as the `write.table( )` function.

```{r, eval=FALSE}
downloadHandler(filename = function() file_name(), # custom filename
                content = function(file){
                  writeLines(file,
                             text = text()) # metadata
                  write.table(population_table(), # population table
                              file = file,
                              sep = ",",
                              append = TRUE,
                              row.names = FALSE,
                              col.names = FALSE,
                              quote = FALSE)})
```

Users can customize the name of the downloaded file.

```{r, eval=FALSE}
file_name <- reactive({
  if (isTruthy(input$filename)) {
    if (input$add_date == TRUE) {paste0(Sys.Date(), "_", input$filename,".csv")}
    else {paste0(input$filename, ".csv", sep = "")}
  }
  else {
    if (input$add_date == TRUE) {paste0(Sys.Date(), "_", "population_statistics.csv")}
    else {paste0("population_statistics.csv")}
  }
})
```

